Optimal (Multiway) Spatial Joins

### 2024-05-28 msj.tex
yufei

Next, we will prove that if the 3D 3-SJ problem can be solved in $O((n + \out) \cdot \polylog n)$ time, then the triangle detection problem can be solved in $O(m \polylog m)$ time. This would be truly surprising because as mentioned in Section~\ref{sec:intro:ours} the state of the art algorithm for triangle detection runs in $O(m^{1.41})$ time \cite{ayz97}. Thus, in the absence of such a breakthrough, no $O((n + \out) \cdot \polylog n)$ time algorithms can exist for the 3D 3-SJ problem.

@@

Given a graph $G = (V,E)$ for triangle detection, we will construct an instance of the 3D 3-SJ problem with input size $3m$. W.l.o.g., let us assume that each vertex of $V$ is represented as a unique integer in $[|V|]$. Initialize 3 empty sets of rectangles: $R_1, R_2$, and $R_3$. For each edge $\set{u, v}\in E$ where $u < v$, we add rectangles
\begin{itemize}
    \item $(-\infty, \infty) \times [u,u] \times [v,v]$ to $R_1$;
    \item $[v,v] \times(-\infty, \infty)\times [u,u]$ to $R_2$;
    \item $[v,v] \times [u,u] \times(-\infty, \infty) $ to $R_3$.
\end{itemize}
Note that every rectangle in $R_1$ (resp., $R_2$ and $R_3$) has $(-\infty, \infty)$ as the x- (resp., y- and z-) projection.

@@

The construction has the property that $G$ has a triangle if and only if $\J(R_1, R_2, R_3) \neq \emptyset$. We can prove this with the following argument.
\myitems{
    \item Suppose that $G$ with triangle with vertices $u$, $v$, and $w$ such that $u < v  < w$. Then, by our construction, $r_1 = (-\infty, \infty) \times [u,u] \times [v,v] \in R_1$, $r_2 = [w, w] \times (-\infty, \infty) \times[v,v] \in R_2$, and $r_3 = [w,w] \times [u, u] \times (-\infty, \infty) \in r_3$. It is clear that $(r_1, r_2, r_3)$ is a result tuple in $\J(R_1, R_2, R_3)$.

@@

Now, assume that there exists an algorithm $\A$ capable of solving the 3D 3-SJ problem in $O((n + \out) \cdot \polylog n)$ time. For $\out = 0$, this algorithm must perform at most $c \cdot n \polylog n$ steps when executed on an input with input size $n$. We run $\A$ on the 3D 3-SJ instance $R_1, R_2, R_3$ constructed earlier in a {\em cost-monitoring manner}:
\myitems{
    \item If $\A$ terminates within $c \cdot (3m) \polylog (3m)$ steps, we check whether it finds any result tuple in $\J(R_1, R_2, R_3)$. If so, a triangle has been found in $G$; otherwise, we declare that $G$ has no triangles.

    \item If $\A$ has performed $1 + c \cdot (3m) \polylog (3m)$ steps, we manually terminate the algorithm and declare that $\J(R_1, R_2, R_3) \ne \empty$, meaning that $G$ must have at least one triangle.
}
The above strategy thus settles the triangle detection problem in $O(m\polylog m)$ time.

>>>>

The subsequent discussion will show that if the 3D 3-SJ problem can be solved in $O((n + \out) \cdot \polylog n)$ time, then the triangle detection problem can be solved in $O(m \polylog m)$ time. This would be truly surprising because as mentioned in Section~\ref{sec:intro:ours} the state-of-the-art algorithm for triangle detection runs in $O(m^{1.41})$ time \cite{ayz97}. Thus, in the absence of such a breakthrough, no $O((n + \out) \cdot \polylog n)$ time algorithms can exist for the 3D 3-SJ problem.

@@

Given a graph $G = (V,E)$ for triangle detection, we will construct an instance of the 3D 3-SJ problem with input size $3m$. W.l.o.g., let us assume that each vertex of $V$ is represented as a unique integer in $[|V|]$. Initialize $R_1, R_2$, and $R_3$ as 3 empty sets of rectangles. For each edge $\set{u, v}\in E$ where $u < v$, we add
\begin{itemize}
    \item a rectangle $(-\infty, \infty) \times [u,u] \times [v,v]$ to $R_1$;
    \item $[v,v] \times(-\infty, \infty)\times [u,u]$ to $R_2$;
    \item $[v,v] \times [u,u] \times(-\infty, \infty) $ to $R_3$.
\end{itemize}
Note that every rectangle in $R_1$ (resp., $R_2$ and $R_3$) has $(-\infty, \infty)$ as the x- (resp., y- and z-) projection.

@@

The construction has the property that $G$ has a triangle if and only if $\J(R_1, R_2, R_3) \neq \emptyset$. We can prove this with the following argument.
\myitems{
    \item Suppose that $G$ has a triangle with vertices $u$, $v$, and $w$ such that $u < v  < w$. Then, by our construction, $r_1 = (-\infty, \infty) \times [u,u] \times [v,v] \in R_1$, $r_2 = [w, w] \times (-\infty, \infty) \times[v,v] \in R_2$, and $r_3 = [w,w] \times [u, u] \times (-\infty, \infty) \in r_3$. It is clear that $(r_1, r_2, r_3)$ is a result tuple in $\J(R_1, R_2, R_3)$.

@@

Now, assume that there exists an algorithm $\A$ capable of solving the 3D 3-SJ problem in $O((n + \out) \cdot \polylog n)$ time. For $\out = 0$, this algorithm must perform at most $c \cdot n \polylog n$ steps when the input size is $n$. We run $\A$ on the 3D 3-SJ instance $R_1, R_2, R_3$ constructed earlier in a {\em cost-monitoring manner}:
\myitems{
    \item If $\A$ terminates within $c \cdot (3m) \polylog (3m)$ steps, we check whether it has output any result tuple in $\J(R_1, R_2, R_3)$. If so, a triangle has been found in $G$; otherwise, we declare that $G$ has no triangles.

    \item If $\A$ has performed $1 + c \cdot (3m) \polylog (3m)$ steps, we manually terminate the algorithm and declare that $\J(R_1, R_2, R_3) \ne \emptyset$, meaning that $G$ must have at least one triangle.
}
The above strategy thus settles the triangle detection problem in $O(m\polylog m)$ time.
